Experiences with Planning for Natural Language Generation

### 2009-06-03 trunk/Papers/ci-spark/experiments.tex
revising desc of my experiments

We then go through a series of experiments in the GIVE domain. We
start with a domain that is essentially Gridworld
(Section~\ref{sec:exper-2:-minim}), and then add positions that are
either not needed (Section~\ref{sec:experiment-3:-give}) or even
inaccessible (Section~\ref{sec:experiment-4:-give}). This series of
experiments is set up to illustrate what we perceive as the main
limitation of current off-the-shelf planners for our applications:
They spend a long time on computing all ground instances, even if most
of these instances are never used in the plan search.

@@

For the first experiment on sentence generation, we focus on the
generation of referring expressions, a problem that is usually handled
by the sentence planner if sentence planning and surface realization
are separated. We consider a series of sentence generation problems
which require the planner to compute a plan representing the sentence
``Mary likes the Adj$_1$ \ldots Adj$_n$ rabbit.''  Each problem
instance assumes a target referent $r$, which is a rabbit, and a
certain number $m$ of further rabbits $r_1,\ldots,r_m$ that are
distinguished by properties $P_1,\ldots,P_n$ with $n \leq m$. The
problem instance is set up such that $r$ has all properties except for
$P_i$ in common with each $r_i$. That is, all $n$ properties are
required to describe $r$ uniquely. The $n$ properties are realized as
$n$ different adjectives, in any order.  This setup allows us to vary
the plan length (a plan with $n$ properties will have length $n+4$)
and the universe size (the universe will contain $m+1$ rabbit
individuals in addition to the differently-typed individuals used to
encode the grammar).

@@

We converted these generation problem instances into planning problem
instances as described in Section~\ref{sec:domains}, and then ran
three different planners on them. We used the off-the-shelf planners
FF 2.3 \citep{HoffmannNebel01} and SGPLAN 5.2.2
\citep{hsu06:_new_featur_in_sgplan_for}, because they were highly
successful in the recent International Planning Competitions, and
support a fragment of PDDL with quantified and conditional effects,
which is necessary in our domain (see
Fig.~\ref{fig:white-rabbit-as-planning}). Common to both of these
planners is that they perform a preprocessing step in which all ground
instances of all predicates and actions are computed.  Furthermore, we
used an ad-hoc implementation of GraphPlan \citep{Blum1997} written in
Java which avoids the grounding preprocessing step and only computes
instances of actions and predicates by need, as they are actually
discovered when the planning graph is built. Our expectation was that
this planner would not be able to compete with FF or SGPLAN in terms
of search efficiency, but we were interested in comparing the
contribution of the prior grounding step on the planner performance.

The results of this experiment are shown in the graph in
Figure~\ref{fig:runtimes-crisp}.\footnote{All runtimes in
  Sections~\ref{sec:exper-1:-sent} and \ref{sec:exper-2:-sent-xtag}
  were measured on an AMD Opteron 8220 CPU running at 2.8 GHz, under
  Linux. FF 2.3 and Metric-FF were recompiled as 64-bit binaries. Java
  programs were executed under Java 1.6.0\_13 in 64-bit mode. They
  were allowed to ``warm up'', i.e. the JVM was given the opportunity
  to just-in-time compile the relevant bytecode by running the planner
  three times and discarding the runtimes before taking the actual
  measurements. All runtimes are averaged over three runs of
  the planners.} The input parameters $(m,n)$ are plotted in
lexicographic order on the horizontal axis, and the runtime is shown
in seconds on the vertical axis. These results reveal a number of
interesting insights. First, FF significantly outperforms SGPLAN in
this domain, sometimes by a factor of 10 or more. Second, FF's runtime
is dominated by its initial grounding step, in which it computes the
ground instances of all predicates and actions used in the planning
problem in order to avoid unnecessary instantiation during search.  In
particular, the ratio of grounding time to total runtime is generally
above 85\%, and rises to above 99\% at $m=11$, which is still a
relatively small universe for this application.\footnote{The
  ``grounding'' time reported here is what FF reports as ``time spent:
  instantiating action templates''.}

In our testing examples, the time spent by FF on grounding is so high
that it is consistently outperformed by our Java implementation of
GraphPlan, which only computes instances by need. Although FF is
consistently much faster as far as pure search time is concerned, our
results indicate that FF's performance is much more sensitive to the
domain size: if we fix $n=1$, FF takes 60 ms to compute a plan at
$m=1$, but 2.4 seconds to compute the same plan at $m=10$. By
comparison, our GraphPlan implementation takes 30 ms at $m=1$ and
still only requires 50 ms at $m=10$. Conversely, GraphPlan's runtime
grows much faster with the plan size (i.e., with growing values of $n$
for a fixed $m$). Larger, but still realistically-sized, instances of
the sentence generation problem are still problematic for all the
planners we tested.

>>>>

We then go through a series of experiments in the GIVE domain. We
start with a domain that is essentially Gridworld
(Section~\ref{sec:exper-2:-minim}), and then add positions that are
either not needed (Section~\ref{sec:experiment-3:-give}) or even
inaccessible (Section~\ref{sec:experiment-4:-give}). This series of
experiments is set up to illustrate what we perceive as the main
limitation of current off-the-shelf planners for our applications:
They spend a long time on computing all ground instances, even if most
of these instances are never used in the plan search. \todo{Ron, this
  is something you need to adapt to the new story -- AK}

@@

For the first experiment on sentence generation, we exercise the
ability of the CRISP system described in
Section~\ref{sec:domain-crisp} to generate referring expressions. This
problem is usually handled by the sentence planner if sentence
planning and surface realization are separated; here it happens as
part of the overall generation process.

We consider a series of sentence generation problems which require the
planner to compute a plan representing the sentence ``Mary likes the
Adj$_1$ \ldots Adj$_n$ rabbit.''  Each problem instance assumes a
target referent $r$, which is a rabbit, and a certain number $m$ of
further rabbits $r_1,\ldots,r_m$ that are distinguished by properties
$P_1,\ldots,P_n$ with $n \leq m$. The problem instance is set up such
that $r$ has all properties except for $P_i$ in common with each $r_i$
for $1 \leq i \leq n$, and $r_{n+1},\ldots,r_m$ have none of the
properties $P_i$. That is, all $n$ properties are required to describe
$r$ uniquely. The $n$ properties are realized as $n$ different
adjectives, in any order.  This setup allows us to vary the plan
length (a plan with $n$ properties will have length $n+4$) and the
universe size (the universe will contain $m+1$ rabbit individuals in
addition to the individuals used to encode the grammar, which have
different types).

@@

We converted these generation problem instances into planning problem
instances as described in Section~\ref{sec:domains}, and then ran
several different planners on them. We used three off-the-shelf
planners: FF 2.3 \citep{HoffmannNebel01}, Metric-FF
\citep{hoffmann:ecai-02}, and SGPLAN 6
\citep{hsu06:_new_featur_in_sgplan_for}; all of these were highly
successful at the recent IPC competitions, and unlike many other IPC
participants, support a fragment of PDDL with quantified and
conditional effects, which is necessary in our domain. In addition, we
used an ad-hoc implementation of GraphPlan \citep{Blum1997} written in
Java; unlike the three off-the-shelf planners, it only computes
instances of literals and operators as they are needed in the course
of the plan search, instead of computing all ground instances in a
separate preprocessing step. Finally, we reimplemented the incomplete
greedy search algorithm used in the SPUD system \citep{Stone2003a}.

The results of this experiment are shown in the graph in
Figure~\ref{fig:runtimes-crisp}.\footnote{All runtimes in
  Sections~\ref{sec:exper-1:-sent} and \ref{sec:exper-2:-sent-xtag}
  were measured on a single core of an AMD Opteron 8220 CPU running at
  2.8 GHz, under Linux. FF 2.3 and Metric-FF were recompiled as 64-bit
  binaries. Java programs were executed under Java 1.6.0\_13 in 64-bit
  mode. They were allowed to ``warm up'', i.e. the JVM was given the
  opportunity to just-in-time compile the relevant bytecode by running
  the planner three times and discarding the runtimes before taking
  the actual measurements. All runtimes are averaged over three runs
  of the planners.} The input parameters $(m,n)$ are plotted in
lexicographic order on the horizontal axis, and the runtime is shown
in seconds on the vertical axis, on a logarithmic scale. These results
reveal a number of interesting insights. First, the search times of FF
and Metric-FF (shown as thinner lines) significantly outperform
SGPLAN's search in this domain---on the largest instances, by a factor
of over 100.\footnote{For FF and Metric-FF, we report the
  ``searching'' and ``total'' times reported by the planners. For
  SGPLAN, we report the ``total'' time and the difference between the
  ``total'' and ``parsing'' times.} Second, FF and Metric-FF perform
very similarly to each other, and their search times are almost the
same as those of the SPUD algorithm, which is impressive because they
are complete search algorithms, whereas SPUD's greedy algorithm is
not.

Finally, it is striking that for all three off-the-shelf planners, the
search only accounts for a tiny fraction of the total runtime; in each
case, the preprocessing times are higher than the search times by one
or two orders of magnitude. As a consequence, even our relatively
naive Java implementation of GraphPlan outperforms them all in terms
of total runtime, because it only computes instances by need. Although
FF is consistently much faster as far as pure search time is
concerned, our results indicate that FF's performance is much more
sensitive to the domain size: if we fix $n=1$, FF takes 27
milliseconds to compute a plan at $m=1$, but 4.4 seconds to compute
the same plan at $m=10$. By comparison, our GraphPlan implementation
takes 20 ms at $m=1$ and still only requires 22 ms at
$m=10$.
